---
title: "Install R Packages"
editor: visual
---

## Packages we will need

We set some options to make sure that we include dependencies and also that we install pre-compiled binaries.

I'm sure we'll have forgotten some, but we'll install them when needed.

We can also install packages by using `Install` button on the **Packages** tab.

```{r}
#| echo: true
#| eval: false
install.packages("tidyverse", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
install.packages("modelr", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
install.packages("readr", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
install.packages("huxtable", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
# To be used to set up git/github
install.packages("usethis", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
install.packages("credentials", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
install.packages("styler", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
# For packages not on CRAN we need devtools
install.packages("devtools", dependencies = TRUE, 
                 type="binary", repos='http://cran.uib.no')
```

I often use the following trick at the end of my documents, a simple R chunk with a single comment.
Clicking the "Run All Chunks Above" above button will run all chunks in the document.

```{r}
#| echo: true
#siste
```
