---
title: "Joins"
author: "ag"
format:
  revealjs:
        theme: simple
        footer: '[MSB105](/index.html)'
        reveal_options:
        incremental: true
        smaller: true
        scrollable: true
        slide-number: c/t
editor: visual
editor_options:
  markdown:
    wrap: 72
    canonical: true
    chunk_output_type: console
echo: true
eval: true
bibliography: ag_joins.bib
nocite: '@*'
---

```{r setup}
library(tidyverse)
library(nycflights13)

options(paged.print = FALSE)
```

```{r bib, include=FALSE}
# create a bib file for the R packages 
# used in this document
# Note! Needs to do a touch(ag_ds.bib) in terminal before first run
# else stops when bibliography: "ag_ds.bib" not found in YAML
knitr::write_bib(c(
  "tidyverse",
  "nycflights13"
),
file = "ag_joins.bib"
)
```

## Relational data (chap. 13)

-   Data spread over files/tables and some `key` existing, or possible
    to generate, that we can use to tie them together

-   **Mutating joins**: add new variable(s) to a tibble/dataframe taken
    from other tibbles/dataframes

-   **Filtering joins**: filter *observations*, hence affect
    observations not variables.

-   **Set observations**: Think Venn-diagrams, union, intersection, set
    differences

-   Package `dplyr`: "simplified" SQL

## Example data

-   package `nycflights13`

    -   airlines

    -   airports

    -   weather

    -   planes

    -   flights

-   `help(flights)` "On-time data for all flights that **departed** NYC"

-   dep_time, arr_time: Actual departure and arrival times (format HHMM
    or HMM), local tz.

## Example data

```{r}
head(airlines, n = 3)
```

```{r}
flights <- nycflights13::flights
flights %>% 
  select(year:day, carrier, everything()) %>% 
  group_by(carrier) %>% 
  summarise(first(dep_time)) %>% 
  head(n= 3)
```

-   **Carrier** possible key to get airlines names into the flights
    dataframe.

## Example data

-   From `help(weather)`: Hourly meterological data for LGA, JFK and
    EWR.

. . .

```{r}
head(weather, n = 3)
```

-   **Origin** possible key to match airport and weather, but not
    enough. We also have to match on year, month, day and hour.

## Example data

```{r}
planes %>%
  head(n = 3)
```

## Example data

```{r}
flights %>% 
  select(tailnum, everything()) %>% 
  head(n = 3)
```

-   Probably enough with `tailnum` as key.

## Example data

```{r}
airports %>% 
  filter(faa %in% c('JFK', 'LGA', 'EWR'))
```

-   faa seems to be the same code we have in `flights$origin`. Check the
    variables or documentation.

## Join flights and airline

-   left_join(): includes all rows in x, see `help(left_join)` for
    alternatives
-   note that *key* is the same (*carrier*) in both tables.

. . .

```{r}
flights <- nycflights13::flights 
flights <- flights%>% 
  # note x = flights, y = airlines
  left_join(airlines, by = join_by("carrier"))

flights %>% 
  select(carrier, name, everything()) %>% 
  head(3)
```

## Join flights and planes

-   check names

. . .

```{r}
nmf <- names(flights)
nmf[nmf %in% names(planes)]
```

. . .

-   simpler solution

. . .

```{r}
# alternative
intersect(names(flights), names(planes))
```

-   "year" has two different meanings, year of the flight and the year
    the plane was built. We have to keep them apart (suffix option).

## Join flights and planes

```{r}
flights <- flights %>% 
  left_join(planes, by = join_by("tailnum"), 
            suffix = c("", "_plane"))

flights %>% 
  select(tailnum, model, 
         year_plane, year, everything()) %>% 
  head(n = 3) |> 
  print(width = 80)
```

## Join origin and dest with airport data

-   `origin` is just the three NYC airports (faa code). Check with
    (`unique(flights$origin)`)
-   `dest` is the destination, about 100 different airports (faa code)
-   `faa` airport code in airports tibble

## Join origin and dest with airport data

-   check names

. . .

```{r}
nmf <- names(flights)
nmf[nmf %in% names(airports)]
```

```{r}
head(flights$name, n = 4) |> 
  print(width = 80)
```

```{r}
head(airports$name, n = 4) |> 
  print(width = 80)
```

## Join origin with airport data

```{r}
#| paged.print: false
flights <- flights %>% 
  left_join(airports,
            # Note new syntax
            by = join_by(origin == faa), 
            suffix = c("_airline", "_origin_airport"), 
            keep = TRUE
            ) %>% 
  select(year:day, name_airline, name_origin_airport, everything())
head(flights, n = 3) |> 
   print(width = 80)
```

## Join dest with airport data

-   Now we have to be really careful with names.

-   check names

. . .

```{r}
nmf <- names(flights)
nmf[nmf %in% names(airports)]
```

-   We are going to use `faa` as key and can set `keep = FALSE` to get
    rid of the one in `airports`.
-   The others we will suffix with "\_origin" and "\_dest".

## Join dest with airport data

```{r}
#work with tmp until correct then 
#change to flights
flights <- flights %>% 
  left_join(airports, 
            by = c("dest" = "faa"),
            suffix = c("_origin", "_dest"),
            keep = FALSE
            ) |> 
  select(faa, origin, dest, lat_origin, lat_dest, everything())

names(flights) |> 
  print(width = 80)
```

## Join origin airport and weather

-   Luckily we only have weather data for the origin (NYC) airport ;-)
-   We want to know the weather conditions on departure
-   Weather has data for every date and every hour, hour 0 to 23.
    -   Not all days have all hours. Probably dropped if no flights.
    -   Probably the reason first row in weather is hour = 1 for EWR.
-   dep_time: Actual departure and arrival times (format HHMM or HMM),
    local tz.
-   We have to get the hour out of dep_time, ex. 513 should be 5 while
    1309 should be 13
-   Many solutions, convert to datetime object etc.
-   We will pad with 0 in front to get 4 characters, then str_sub the
    two first characters and convert to integer.

## Check hour var. in weather

```{r}
weather |> filter(origin == "EWR") |> select(hour) |> table()
```

```{r}
#| code-fold: false
weather |> filter(origin == "LGA") |> select(hour) |> table()
```

```{r}
#| code-fold: false
weather |> filter(origin == "LGA") |> select(hour) |> table()
```

-   This means we just have to get the two leftmost characters of
    dep_time to get the hour.

## Join origin airport and weather, fix dep_time

-   We first develop our solution with less data. No need to use
    `r dim(flights)[1]` observations while we develop our solution.

. . .

```{r}
set.seed(666)
(x <- sample(flights$dep_time, size = 30)) |> 
  print(width = 80)
```

-   Good mix, NA, 29, 826, 1156

## Join origin airport and weather, fix dep_time

```{r}
x <- str_pad(x, width = 4, side = "left", pad = "0")
x |> print(width = 80)
```

## Join origin airport and weather, fix dep_time

```{r}
(x <- as.numeric(str_sub(x, start = 1, end = 2))) |> 
  print(width = 80)
```

## Join origin airport and weather; fix_dep_time

```{r fix_dep_time}
fix_dep_time <- function(x) {
  x = str_pad(x, width = 4, side = "left", pad = "0")
  x = str_sub(x, start = 1, end = 2)
  # This is correct since hour 0: 23 in weather dataset
  # Note that not all days have hour = 0
  x = as.integer(x)
  return(x)
  }
```

## Join origin airport and weather; dep_time_hour

```{r}
flights <- flights %>% 
  mutate(
    dep_time_hour = fix_dep_time(dep_time)
  )
```

## Join origin airport and weather

-   check names

. . .

```{r}
nmf <- names(flights)
nmf[nmf %in% names(weather)]
```

-   first 5 we use as key, drop "y version" with `keep = FALSE`. Change
    name om time_hour to time_hour_weather with suffix.

## Join origin airport and weather

```{r}
flights <- flights %>% 
  left_join(weather, 
            #by = c("origin", "year", "month", "day", "dep_time_hour" = "hour"),
            by = join_by(origin, year, month, day, dep_time_hour == hour),
            suffix = c("", "_weather"), 
            keep = FALSE
            )
```

-   We also ought to group the variables a bit better with `select()`.

. . .

```{r}
#| paged.print: false
flights <- flights %>% 
  select(year, month, day, dep_time, everything())

flights %>%  print(width = 80)
```

```{r}
flights[1:300,] |> View()
```

## The other joins

-   "The left join should be your default join
    -   use it unless you have a strong reason to prefer one of the
        others."
-   If you *really* think you need them, you will find them all in Chap.
    13

## References
